---
title: "Advent of Code: Day 20+"
output: html_notebook
---

# Day 20
```{r}
require(tidyverse)
algorithm <- scan('data/aoc-20_algorithm.txt', character())
a_char <- str_split(algorithm, '')[[1]]
alg <- if_else(a_char == ".", 0, 1)
d <- read_csv('data/aoc-20_input.txt', col_names=c('dat'),
              col_types=cols(.default = col_character()))
d <- d %>%
  separate(dat, into=str_c('c', 1:100), sep=1:100)
dm <- d %>%
  mutate_all(function(x) {
    if_else(x==".",0,1)
    }) %>%
  as.matrix(nrow=100)
```

Borrow `bin2dec()` from Day 16.
```{r}
bin2dec <- function(b) {
  n <- length(b)
  d <- 0
  for (ii in 1:n) {
    d <- d + as.numeric(b[n-ii+1])*(2^(ii-1))
  }
  return(d)
}

score_position <- function(mat, r, c) {
  cols <- (c-1):(c+1)
  r1 <- mat[ r-1, cols]
  r2 <- mat[ r  , cols]
  r3 <- mat[ r+1, cols]
  val <- bin2dec(c(r1, r2, r3))
  return(val)
}

enhance <- function(dm, iteration=1) {
  
  n_row <- nrow(dm)
  n_col <- ncol(dm)
  pad <- 4
  pad_offset <- pad/2
  # expand in each direction and pad with 0s
  # except that won't work because 9 0s -> 1
  # we need to know the iteration to know whether to pad with 1s or 0s
  if (iteration %% 2 == 0) {
    out_mat <- matrix(1, n_row+pad, n_col+pad)
  } else {
    out_mat <- matrix(0, n_row+pad, n_col+pad)
  }
  in_mat <- out_mat
  in_mat[pad_offset:(n_row+pad_offset-1) , pad_offset:(n_col+pad_offset-1)] <- dm[1:n_row, 1:n_col]
  
  out_row <- nrow(out_mat)
  out_col <- ncol(out_mat)
  # use a moving 3x3 window to get output at each input point
  for (r in 2:(nrow(in_mat)-1) ) {
    for (c in 2:(ncol(in_mat)-1) ) {
      v <- score_position(in_mat, r, c)
      out_mat[r, c] <- alg[v+1]
    }
  }
  # handle outer edge
  out_mat[c(1, out_row), 1:out_col] <- iteration %% 2
  out_mat[1:out_row, c(1, out_col)] <- iteration %% 2

  return(out_mat)
}
```

```{r}
test <- matrix(0, 15, 15)
test[6,6] <- 1
test[6,9] <- 1
test[7,6] <- 1
test[8,6] <- 1
test[8,7] <- 1
test[8,11] <- 1
test[9,8] <- 1
test[10,8] <- 1
test[10,9] <- 1
test[10,10] <- 1
```


```{r}
dm %>% enhance() %>% enhance(2) %>% sum()
```

## Part 2
```{r}
dmw <- dm
iter <- 1
for (i in 1:50) {
  dmw <- enhance(dmw, iter)
  iter <- iter+1
}
```

```{r}
dmw %>% sum()
```

# Day 21: Dirac Dice

A die rolls deterministically starting at 1, incrementing to 100, then circling back to 1.

We'll make some functions to simulate three rolls of the dice
```{r}
require(tidyverse)
die <- list('last'=100, 'vals'=c(0,0,0), 'rolls'=0)
class(die) <- 'Die'

roll <- function(object, ...) UseMethod('roll')

roll.Die <- function(object, last) {
  if (missing(last)) {
    last <- object$last
  }
  last <- last + 1
  r <- last:(last+2)
  r[r > 100] <- r[r > 100] - 100
  object$vals <- r
  object$last <- r[3]
  object$rolls <- object$rolls + 3
  return(object)
}

player <- list('start'=1, 'position'=1, 'score'=0)
class(player) <- 'Player'


init <- function(object, ...) UseMethod('init')
move <- function(object, ...) UseMethod('move')

init.Player <- function(object, start=1) {
  object$start <- start
  object$position <- start
  object$score <- 0
  return(object)
}

move.Player <- function(object, d=die) {
  # get rolls
  d <- roll(d)
  rsum <- sum(d$vals)
  # shift position
  pos <- object$position + rsum
  if (pos > 10) {
    pos <- pos %% 10
    if (pos == 0) pos <- 10
  }
  object$position <- pos
  object$score <- object$score + pos
  out <- list('player'=object, 'die'=d)
  return(out)
}
```

```{r}
d <- die
player1 <- init(player, start=7)
player2 <- init(player, start=3)
```

```{r}
winner <- FALSE
while (!winner) {
  res <- move(player1, d)
  player1 <- res$player
  d <- res$die
  if (player1$score > 999) {
    winner <- TRUE
    print('Player 1 wins!')
  } else {
    res <- move(player2, d)
    d <- res$die
    player2 <- res$player
  }
  if (player2$score > 999) {
    winner <- TRUE
    print('Player 2 wins!')
  } 
}
```

```{r}
player2$score * d$rolls
```

## Part 2: Dirac dice
Now the die has 3 sides. We play until one player reaches 21. Of all possible games, how many games does Player 1 win and how many does player 2 win.

Hmmm...that's a lot of possibilities. Let's start with how many different results we could get from 3 rolls
* 3 : 1 (111)
* 4 : 3 (112, 121, 211)
* 5 : 4 (122, 212, 221, 113, 131, 331)
* 6 : 7 (222, 123, 321, 213, 231, 312, 132)
* 7 : 4 (322, 232, 233, 133, 313, 311)
* 8 : 3 (332, 323, 233)
* 9 : 1 (333)

So, of the 27 possible rolls, there are really only 7 possible outcomes

```{r}
possibilities <- c(3:9)
results1 <- tibble(
  r1 = rep(possibilities, 7*7),
  r2 = rep(possibilities, each=7, 7),
  r3 = rep(possibilities, each=7*7)
)
results1$start <- 7
results1 <- results1 %>%
  mutate(p1 = case_when(
    r1 + start > 10 ~ r1 + start - 10,
    TRUE ~ r1 + start
  )) %>% 
  mutate(p2 = case_when(
    p1 + r2 > 10 ~ p1 + r2 - 10,
    TRUE ~ p1 + r2
  )) %>%
  mutate(p3 = case_when(
    p2 + r3 > 10 ~ p2 + r3 - 10,
    TRUE ~ p2 + r3
  )) %>%
  mutate(score = p1+p2+p3)

results1 %>% count(score)
```

```{r}
results2 <- tibble(
  r1 = rep(possibilities, 7*7),
  r2 = rep(possibilities, each=7, 7),
  r3 = rep(possibilities, each=7*7)
)
results2$start <- 3
results2 <- results2 %>%
  mutate(p1 = case_when(
    r1 + start > 10 ~ r1 + start - 10,
    TRUE ~ r1 + start
  )) %>% 
  mutate(p2 = case_when(
    p1 + r2 > 10 ~ p1 + r2 - 10,
    TRUE ~ p1 + r2
  )) %>%
  mutate(p3 = case_when(
    p2 + r3 > 10 ~ p2 + r3 - 10,
    TRUE ~ p2 + r3
  )) %>%
  mutate(score = p1+p2+p3)

results2 %>% count(score)
```

```{r}
comp <- tibble(score = c(results1$score, results2$score),
               player = rep(c('1', '2'), each=nrow(results1)))
comp %>%
  ggplot(aes(score, color=player, fill=player)) + 
  geom_density(alpha=.3) +
  theme_classic()
```



```{r}

move_Dirac <- function(object, ...) UseMethod('move_Dirac')
move_Dirac.Player <- function(object, nums=c(3,3,3)) {
  rsum <- sum(nums)
  # shift position
  pos <- object$position + rsum
  if (pos > 10) {
    pos <- pos %% 10
    if (pos == 0) pos <- 10
  }
  object$position <- pos
  object$score <- object$score + pos
  return(object)
}

play_game <- function(player1, player2, rolls) {
  
  roll_idx <- 1
  winner <- FALSE
  while (!winner) {
    r <- rolls[roll_idx,]
    player1 <- move_Dirac(player1, r)
    roll_idx <- roll_idx+1
    if (roll_idx > nrow(rolls)) {
      roll_idx <- 1
      }
    if (player1$score > 20) {
      winner <- TRUE
      print('Player 1 wins!')
    } else {
      r <- rolls[roll_idx,]
      player2 <- move_Dirac(player2, r)
      roll_idx <- roll_idx+1
      if (roll_idx > nrow(rolls)) {
        roll_idx <- 1
      }
    }
    if (player2$score > 20) {
      winner <- TRUE
      print('Player 2 wins!')
    } 
  }
  players <- c(player1, player2)
  return(players)
}
```

```{r}
player1 <- init(player, start=7)
player2 <- init(player, start=3)
```

```{r}
g <- play_game(player1, player2, matrix(3, nrow=1, ncol=3))
```

```{r}
# Solution from https://github.com/AdroMine/AdventOfCode/blob/main/2021/Day21/solution.R
# Brute force option to iterate over all games, but create a dictionary 'cache'
# to store calls to save time and not have to make new calcaultions for duplciate states

# sample
p1 <- 7 ; p2 <- 3

# memoise fails to do anything, maybe using it incorrectly? even after setting
# cachem manually to Inf, it still failed to run in proper time
cache <- collections::dict()

die_sums <- rowSums(expand.grid(1:3, 1:3, 1:3))
die_rolls <- table(die_sums)
die_sums <- unique(die_sums)

game <- function(p1, p2, s1, s2){
    
    if(s1 >= 21){
        return(c(1,0))
    } 
    if(s2 >= 21){
        return(c(0,1))
    }
    key <- paste(p1, p2, s1, s2, sep = ":")
    if(cache$has(key)){
        return(cache$get(key))
    }
    
    res <- c(0, 0)
    
    for(die in die_sums){
        
        new_pos <- ((p1 - 1) + die) %% 10 + 1
        new_sc <- s1 + new_pos
        k <-  rev(game(p2, new_pos, s2, new_sc)) * die_rolls[as.character(die)]
        res <- res + k
    }
    cache$set(key, res)
    res
}

as.character(max(game(p1, p2, 0, 0)))
```
